Analysis paper

### 2021-05-17 analysis/analysis.Rmd
update report and move data wrangling into package

```{r helper-functions}
# classify epidemic according to whether, on a given forecast date, the two
# weeks before that have seen monotonic rise, decline, or an unclear trend
classify_epidemic <- function(data, cutoff = 0.05, growth_cutoff = 2) {
  dt <- as.data.table(data)

  # calculate differences and set differences smaller than 
  # a certain cutoff to zero
  dt[, diff_1 := c(NA, diff(true_value, 1)), 
     by = c("location_name", "target_type")]
  dt[abs(diff_1) < (cutoff * true_value), diff_1 := 0]
  dt[, diff_prev := c(shift(diff_1, 1)), 
     by = c("location_name", "target_type")]

  # assign a label depending on observed differences
  dt[, c("classification", "speed") := "unclear"]
  dt[(diff_1 >= 0 & diff_prev >= 0), 
     classification := "increasing"]
  dt[(diff_1 <= 0 & diff_prev <= 0 ), 
     classification := "decreasing"]
  dt[(diff_1 == 0 & diff_prev == 0), 
     classification := "unclear"]
  dt[(classification == "increasing") & 
       diff_1 > growth_cutoff * diff_prev, 
     speed := "accelerating"]
    dt[(classification == "increasing") & 
         diff_1 < 1/growth_cutoff * diff_prev, 
     speed := "decelerating"]
  dt[(classification == "decreasing") & 
       abs(diff_1) > growth_cutoff * abs(diff_prev), 
     speed := "accelerating"]
  dt[(classification == "decreasing") & 
       abs(diff_1) < 1/growth_cutoff * abs(diff_prev), 
     speed := "decelerating"]

  dt[, c("diff_1", "diff_prev") := NULL]
  return(dt)
}

```{r}
# load and set up data
cases <- combined_data[target_type == "case" & 
                         as.Date(forecast_date) %in% forecast_dates$cases]
deaths <- combined_data[target_type == "death" & 
                         as.Date(forecast_date) %in% forecast_dates$deaths]
data <- rbindlist(list(cases, deaths))
data[, location_target := paste0(target_type, "s", " in ", location_name)]

# obtain classification
epitrend <- classify_epidemic(truth_data)
epitrend[, forecast_date := target_end_date + 2]
epitrend[, true_value := NULL]

@@

ggsave(here("analysis", "plots", "daily_truth.png"))
```
## Number of expert forecasters

@@

```
```{r, results = 'asis'}
dt <- crowdforecast_data[!(model %in% c("Crowd-Rt-Forecast",
                                        "EpiNow2_secondary", 
                                        "EpiExpert-ensemble", 
                                        "EpiNow2")) & forecast_date %in% forecast_dates$unfiltered]

@@

@@

```{r truth-vs-forecasts, fig.height=10, fig.width=15}
# load unfiltered combined data 
df <- combined_data[as.Date(forecast_date) %in% forecast_dates$unfiltered &
                      grepl("2", target) & 
                      quantile %in% c(0.025, 0.25, 0.75, 0.975)]
df[, location_target := str_to_title(paste0(target_type, "s", " in ", location_name))]

df <- dcast(df, ... ~ quantile, value.var = "prediction")

# add trend classification
df <- merge(df, epitrend, 
            by = c("target_end_date", "location_name", "target_type"))

truth_vs_forecast_2 <- df %>%
  ggplot(aes(y = true_value, x = target_end_date)) + 
  # geom_linerange(aes(ymin = `0.025`, ymax = `0.975`, color = model), 
  #                size = 1.5,
  #                alpha = 0.4,
  #                position = position_dodge(width = 3)) + 
  geom_linerange(aes(ymin = `0.25`, ymax = `0.75`, color = model), 
                 size = 1.5,
                 alpha = 1,
                 position = position_dodge(width = 6)) + 
  labs(y = "value", x = "date") + 
  geom_point(aes(shape = classification)) + 
  geom_line() + 
  geom_rect(aes(xmin = min(forecast_dates$christmas) + 14, # 2wk ahead forecast
                xmax = max(forecast_dates$christmas) + 14,
                ymin = -Inf,
                ymax = Inf), fill = "lightgrey", alpha = 0.002) +
  theme_light() + 
  facet_wrap(~ location_target, scales = "free", ncol = 1) + 
  theme(legend.position = "bottom") + 
  scale_color_brewer(palette = "Accent") + 
  theme(panel.background = element_rect(fill = "#e1f0fe"))

ggsave(here("analysis", "plots", "truth-and-2-wk-ahead.png"), 
       plot = truth_vs_forecast_2)

```{r scores-over-time, fig.height=10, fig.width=15}
# plot with scores over time ---------------------------------------------------
scores_target_forecastdate <- eval_forecasts(
  data = data[grepl("2", target)],
  summarise_by = c("model", "location_name", "target_type", "forecast_date", "location_target"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)

# add epi trend classification
scores_target_forecastdate[, forecast_date := as.Date(forecast_date)]
scores_target_forecastdate <-  merge(scores_target_forecastdate, epitrend, 
            by = c("forecast_date", "location_name", "target_type"))

rel_skill_time_2 <- scores_target_forecastdate  %>%
  ggplot(aes(x = as.Date(forecast_date), y = scaled_rel_skill)) + 
  geom_line(aes(colour = model)) + 
  geom_point(aes(shape = classification, colour = model)) + 
  theme_light() + 
  facet_wrap( ~ location_target, ncol = 1, scales = "free_y") +
  theme(legend.position = "bottom") +
  labs(y = "relative skill", x = "date") + 
  geom_rect(aes(xmin = min(forecast_dates$christmas) + 14,
                xmax = max(forecast_dates$christmas) + 14,
                ymin = 0,
                ymax = Inf), fill = "lightgrey", alpha = 0.002) +
  scale_color_brewer(palette = "Accent") + 
  theme(panel.background = element_rect(fill = "#e1f0fe")) +
  scale_y_log10()

ggsave(here("analysis", "plots", "rel-skill-over-time.png"), 
       plot = rel_skill_time_2)
```

```{r fig.height=10, fig.width=15}
# glue two plots together
truth_vs_forecast_2 | rel_skill_time_2 +
  plot_layout(guides = "collect") & 
  plot_annotation(tag_levels = 'A') &
  theme(legend.position = 'bottom')

ggsave(here("analysis", "plots", "figure-forecasts-2.png"))
```
*Figure 1. Left: Visualisation of 2 week ahead forecasts against the true observed values. The shape indicates whether there has been a monotonic increase or decrease over the last two weeks, or an unclear trend. Forecasts made during the Christmas period (from 2020-12-19 until 2021-01-02) are grayed out. Right: Corresponding scaled relative skill scores for the forecasts shown on the left. Scaled relative skill scores can be thought of as 'improvement over the baseline model' (see Methods for details).*

---

### 2 weeks ahead
```{r }
# scores by target type for 2 week ahead forecasts

coverage <- eval_forecasts(
  data[grepl("2", target)], 
  summarise_by = c("model", "target_type", "range"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)[range %in% c(50, 90), 
                   .(model, target_type, coverage, range)] %>%
  dcast(formula = ... ~ range, value.var = "coverage")

table <- eval_forecasts(
  data[grepl("2", target)], 
  summarise_by = c("model", "target_type"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
) %>%
  merge(coverage, all.x = TRUE, by = c("model", "target_type")) %>%
  make_score_table()

---

```{r}
# scores by target type for 4 week ahead forecasts
scores <- eval_forecasts(
  data[grepl("2", target)], 
  summarise_by = c("model", "target_type", "location_name", "forecast_date", 
                   "location_target"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)[model != "Baseline"]

scores[, forecast_date := as.Date(forecast_date)]
scores <-  merge(scores, epitrend, 
            by = c("forecast_date", "location_name", "target_type"))

---

## WIS contributions
```{r}
df <- data[grepl("2", target)] %>%
  merge(epitrend, 
        by = c("target_end_date", "location_name", "target_type"))

scores_target <- eval_forecasts(
  df, 
  summarise_by = c("model", "target_type", "classification"), 
  compute_relative_skill = TRUE
)

---

```{r, eval = TRUE}
scores <- eval_forecasts(
  data[, horizon := substring(target, 1, 1)], 
  summarise_by = c("model", "target_type", "location_name", "horizon"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)

```{r, eval = TRUE}
scores_full <- eval_forecasts(
  data[, horizon := substring(target, 1, 1)], 
  summarise_by = c("model", "location_target", "horizon", "forecast_date"), 
  compute_relative_skill = FALSE, 
  baseline = "Baseline"
)

---

```{r, eval = TRUE}
df <- data[grepl("2", target)] %>%
  merge(epitrend, 
        by = c("target_end_date", "location_name", "target_type"))

scores_target <- eval_forecasts(
  df, 
  summarise_by = c("model", "target_type", "classification", "range", "quantile"), 
  compute_relative_skill = FALSE, 
  pit_plots = TRUE
)

  *Figure 4: Comparison of nominal and empirical coverage (percentage of true observed values covered by a given prediction interval). The one thing I find interesting about this plot is that models are apparently much too confident when there is an increasing trend. I feel this is probably actionable advice (even though it could probably be made better using a separate analysis about optimal uncertainty that is beyond the scope of this paper). *

---

- look at first period, second period etc
```{r, eval = FALSE}
scores <- list()
# scores for restricted data set
scores_full <- eval_forecasts(data, 
                              summarise_by = c("model", "target_type", "target"), 
                              compute_relative_skill = TRUE, 
                              baseline = "Baseline")

<!-- - maybe look at correlation between the number of weekly cases and the score? -->
<!-- - maybe it would be interesting to look at other baseline models, e.g. "how hard is it to beat a model that just always continues the trend?" -->
<!-- - look at crowd performance as a function of the number of forecasts --> maybe leave that to a second paper -->

>>>>

@@

ggsave(here("analysis", "plots", "daily_truth.png"))
```

Todo:

-   do we want to keep the decreasing / increasing points at all on the plot?
-   it is a bit odd that here we look at the classification at the target end date, while below we look at it in terms of the forecast_date

```{r truth-vs-forecasts, fig.height=10, fig.width=15}
# load unfiltered combined data 
df <- unfiltered_data[grepl("2", target) & 
                      quantile %in% c(0.025, 0.25, 0.75, 0.975)]
df <- dcast(df, ... ~ quantile, value.var = "prediction")

# add trend classification by target_end_date (and remove classification by forecast_date)
df <- merge(
  df[, c("classification", "speed") := NULL], 
  copy(epitrend)[, forecast_date := NULL], 
  by = c("target_end_date", "location_name", "target_type")
)

truth_and_obs_plot <- function(df) {
  p <- df %>%
  ggplot(aes(y = true_value, x = target_end_date)) + 
  geom_linerange(aes(ymin = `0.25`, ymax = `0.75`, color = model), 
                 size = 1.5,
                 alpha = 1,
                 position = position_dodge(width = 6)) + 
  labs(y = "value", x = "date") + 
  geom_point(aes(shape = classification)) + 
  geom_line() + 
  geom_rect(aes(xmin = min(forecast_dates$christmas) + 14, # 2wk ahead forecast
                xmax = max(forecast_dates$christmas) + 14,
                ymin = -Inf,
                ymax = Inf), fill = "lightgrey", alpha = 0.002) +
  theme_light() + 
  facet_wrap(~ location_target, scales = "free", ncol = 2) + 
  theme(legend.position = "bottom") + 
  theme(panel.background = element_rect(fill = "#e1f0fe")) + 
  scale_x_date(limits = c(min(forecast_dates$full_hub_period) + 14, 
                          max(forecast_dates$full_hub_period)))
  return(p)
}

truth_vs_forecast_2_cases <- df %>%
  dplyr::filter(target_type == "case") %>%
  truth_and_obs_plot() + 
  scale_color_manual(values = brewer.pal(name = "Accent", n = 5)[2:5])

ggsave(here("analysis", "plots", "truth-and-2-wk-ahead_cases.png"), 
       plot = truth_vs_forecast_2_cases)

truth_vs_forecast_2_deaths <- df %>%
  dplyr::filter(target_type == "death") %>%
  truth_and_obs_plot()  + 
  scale_color_manual(values = brewer.pal(name = "Accent", n = 5)[1:5])

ggsave(here("analysis", "plots", "truth-and-2-wk-ahead_deaths.png"), 
       plot = truth_vs_forecast_2_deaths)

```{r scores-over-time, fig.height=10, fig.width=15}
# load unfiltered combined data 
df <- unfiltered_data[grepl("2", target)]

# plot with scores over time ---------------------------------------------------
scores_target_forecastdate <- eval_forecasts(
  data = df,
  summarise_by = c("model", "location_name", "target_type", 
                   "forecast_date", "target_end_date", "location_target", 
                   "classification"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)

scores_over_time_plot <- function(scores) {
  p <- scores %>%
  ggplot(aes(x = as.Date(target_end_date), y = scaled_rel_skill)) + 
  geom_line(aes(colour = model)) + 
  geom_point(aes(shape = classification, colour = model)) + 
  theme_light() + 
  facet_wrap( ~ location_target, ncol = 2, scales = "free_y") +
  theme(legend.position = "bottom") +
  labs(y = "relative skill", x = "date") + 
  geom_rect(aes(xmin = min(forecast_dates$christmas) + 14,
                xmax = max(forecast_dates$christmas) + 14,
                ymin = 0,
                ymax = Inf), fill = "lightgrey", alpha = 0.002) +
  theme(panel.background = element_rect(fill = "#e1f0fe")) +
  scale_y_log10() + 
  scale_x_date(limits = c(min(forecast_dates$full_hub_period) + 14, 
                          max(forecast_dates$full_hub_period)))
  return(p)
}

rel_skill_time_2_case <- scores_target_forecastdate %>%
  dplyr::filter(target_type == "case") %>%
  scores_over_time_plot() + 
  scale_color_manual(values = brewer.pal(name = "Accent", n = 5)[2:5])

ggsave(here("analysis", "plots", "rel-skill-over-time-cases.png"), 
       plot = rel_skill_time_2_case)

rel_skill_time_2_deaths <- scores_target_forecastdate %>%
  dplyr::filter(target_type == "death") %>%
  scores_over_time_plot() + 
  scale_color_brewer(palette = "Accent")

ggsave(here("analysis", "plots", "rel-skill-over-time-deaths.png"), 
       plot = rel_skill_time_2_deaths)
```

```{r fig.height=10, fig.width=15}
# glue two plots together

# cases_plot <- truth_vs_forecast_2_cases / rel_skill_time_2_case + 
#   plot_layout(guides = "collect") & 
#   theme(legend.position = 'bottom') 
# 
# deaths_plot <- truth_vs_forecast_2_deaths / rel_skill_time_2_deaths + 
#   plot_layout(guides = "collect") & 
#   theme(legend.position = 'bottom')

truth_vs_forecast_2_cases / rel_skill_time_2_case / truth_vs_forecast_2_deaths / rel_skill_time_2_deaths + 
  plot_layout(guides = "collect") & 
  plot_annotation(tag_levels = 'A') &
  theme(legend.position = 'bottom')

ggsave(here("analysis", "plots", "figure-forecasts-2.png"))
```

*Figure 1. A and C: Visualisation of 2 week ahead forecasts against the true observed values. The shape indicates whether there has been a monotonic increase or decrease over the last two weeks, or an unclear trend. Forecasts made during the Christmas period (from 2020-12-19 until 2021-01-02) are grayed out. B and D: Corresponding scaled relative skill scores for the forecasts shown on the left. Scaled relative skill scores can be thought of as 'improvement over the baseline model' (see Methods for details).*

------------------------------------------------------------------------

### 2 weeks ahead

```{r }
# scores by target type for 2 week ahead forecasts

coverage <- eval_forecasts(
  filtered_data[grepl("2", target)], 
  summarise_by = c("model", "target_type", "range"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)[range %in% c(50, 90), 
                   .(model, target_type, coverage, range)] %>%
  dcast(formula = ... ~ range, value.var = "coverage")

table <- eval_forecasts(
  filtered_data[grepl("2", target)], 
  summarise_by = c("model", "target_type"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
) %>%
  merge(coverage, all.x = TRUE, by = c("model", "target_type")) %>%
  make_score_table()

------------------------------------------------------------------------

```{r}
# scores by target type for 2 week ahead forecasts
scores <- eval_forecasts(
  filtered_data[grepl("2", target)], 
  summarise_by = c("model", "target_type", "location_name", "forecast_date", 
                   "location_target", "classification"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)[model != "Baseline"]

# scores[, forecast_date := as.character(forecast_date)]
# scores <-  merge(scores, epitrend, 
#             by = c("forecast_date", "location_name", "target_type"))

------------------------------------------------------------------------

## WIS contributions

```{r}
# df <- data[grepl("2", target)] %>%
#   merge(epitrend, 
#         by = c("target_end_date", "location_name", "target_type"))

scores_target <- eval_forecasts(
  filtered_data, 
  summarise_by = c("model", "target_type", "classification"), 
  compute_relative_skill = TRUE
)

------------------------------------------------------------------------

```{r, eval = TRUE}
scores <- eval_forecasts(
  filtered_data, 
  summarise_by = c("model", "target_type", "location_name", "horizon"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)

```{r, eval = TRUE}
scores_full <- eval_forecasts(
  filtered_data, 
  summarise_by = c("model", "location_target", "horizon", "forecast_date"), 
  compute_relative_skill = FALSE, 
  baseline = "Baseline"
)

------------------------------------------------------------------------

```{r, eval = TRUE}
# df <- data[grepl("2", target)] %>%
#   merge(epitrend, 
#         by = c("target_end_date", "location_name", "target_type"))

scores_target <- eval_forecasts(
  filtered_data, 
  summarise_by = c("model", "target_type", "classification", "range", "quantile"), 
  compute_relative_skill = FALSE, 
  pit_plots = TRUE
)

*Figure 4: Comparison of nominal and empirical coverage (percentage of true observed values covered by a given prediction interval). The one thing I find interesting about this plot is that models are apparently much too confident when there is an increasing trend. I feel this is probably actionable advice (even though it could probably be made better using a separate analysis about optimal uncertainty that is beyond the scope of this paper).*

------------------------------------------------------------------------

-   look at first period, second period etc

```{r, eval = FALSE}
scores <- list()
# scores for restricted data set
scores_full <- eval_forecasts(filtered_data, 
                              summarise_by = c("model", "target_type", "target"), 
                              compute_relative_skill = TRUE, 
                              baseline = "Baseline")

<!-- - maybe look at correlation between the number of weekly cases and the score? -->

<!-- - maybe it would be interesting to look at other baseline models, e.g. "how hard is it to beat a model that just always continues the trend?" -->

<!-- - look at crowd performance as a function of the number of forecasts -->

maybe leave that to a second paper --\>

@@

```{r num-forecasters2, results = 'asis'}
dt <- crowdforecast_data[!(model %in% c("Crowd-Rt-Forecast",
                                        "EpiNow2_secondary", 
                                        "EpiExpert-ensemble", 
                                        "EpiNow2")), 
                         .(`number of forecasters` = length(unique(model))), , 
                         by = c("forecast_date", "location_name", "target_type")
][order(forecast_date)][
  !is.na(forecast_date)]

dt %>%
  ggplot(aes(y = `number of forecasters`, x = forecast_date)) + 
  geom_line() + 
  facet_wrap(location_name ~ target_type)

```{r, results = 'asis'}
dt <- crowdforecast_data[!(model %in% c("Crowd-Rt-Forecast",
                                        "EpiNow2_secondary", 
                                        "EpiExpert-ensemble", 
                                        "EpiNow2")) & forecast_date %in% forecast_dates$unfiltered]

@@

median(num_fc$n_forecasts)
mean(num_fc$n_forecasts)

@@
